---
title: "Analysis of pre-polypectomy results"
author: "Yuriy Sverchkov"
date: "4/18/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
library("tidyr")
library("knitr")

age.bin.list = c( "Under 50", "50-54", "55-59", "60-64", "65-9", "70+" )
polyp.locations = c( "Colon", "Colon and Rectum", "Rectum" )
```

```{r prepare data}
summarized.df = readRDS("../data-clean/summarized.v2.rds")

pre.df = summarized.df %>%
  filter( grepl( "Pre$", Sample.ID ) ) %>%
  mutate( `Log-measurement` = log( Geom.Mean ) ) %>%
  rename( `Risk category` = Risk.category.Cancer.stage )

proteins = unique( pre.df$Protein )
n.proteins = length( proteins )
```

Protein measurements (the logarithm of the geometric mean (over replicates) of area ratios):
```{r plot pre vs post, fig.height=10}
ggplot( pre.df, aes( x=Protein, y=`Log-measurement` ) ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

```{r constants for tests}
alpha = 0.05
p.value.threshold = alpha/n.proteins
```

## Comparing high-risk vs. low-risk

```{r low-high counts}
kable( pre.df %>%
         group_by( Protein, `Risk category` ) %>%
         summarise( Count = n() ) %>% ungroup() %>%
         spread( `Risk category`, Count ), caption = "Counts" )
```

Since we test `r n.proteins` proteins, we use a Bonferroni correction and consider only p-values below $`r alpha` / `r n.proteins` = `r p.value.threshold`$ significant.

```{r plot differences by risk}
ggplot( pre.df, aes( x="", y=`Log-measurement`, fill = `Risk category` ) ) +
  facet_wrap( ~Protein, ncol = 3, scales = "free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

Wilcoxon test to determine whether levels of the high-risk group are significantly different from levels of the low-risk group:
```{r difference wilcox test}
wilcox.result = Reduce( rbind.data.frame, Map( function (protein) {
  local.df = pre.df %>% filter( Protein == protein )
  low.df = local.df %>% filter( `Risk category` == "Low" )
  high.df = local.df %>% filter( `Risk category` == "High" )
  result = wilcox.test( x = low.df$`Log-measurement`, y = high.df$`Log-measurement` )
  data.frame( Protein = protein, `p-value` = result$p.value, `Wilcoxon statistic` = result$statistic )
}, proteins) )

rownames( wilcox.result ) = NULL

kable( wilcox.result, caption = "Two-sided Wilcoxon rank sum test" )
```


### Breakdown by gender

```{r pre, lov vs high by gender, fig.height=10}
gender.df = pre.df %>% filter( Gender %in% c("M","F") )

kable( gender.df %>%
         filter( Protein == "APCS" ) %>%
         group_by( `Risk category`, Gender ) %>%
         summarise( Count = n() ) %>% ungroup() %>%
         spread( `Risk category`, Count ), caption = "Counts" )

ggplot( gender.df, aes( x=Gender, y=`Log-measurement`, fill = `Risk category` ) ) +
  facet_wrap( ~Protein, ncol = 3, scales = "free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

### Breakdown by age

```{r plot pre vs post by age, fig.height=10}
age.df = pre.df %>% filter( Age.Bin %in% age.bin.list )

kable( age.df %>%
         filter( Protein == "APCS" ) %>%
         group_by( `Risk category`, Age.Bin ) %>%
         summarise( Count = n() ) %>% ungroup() %>%
         spread( `Risk category`, Count ), caption = "Counts" )

ggplot( age.df, aes( x=Age.Bin, y=`Log-measurement`, fill = `Risk category` ) ) +
  facet_wrap( ~Protein, ncol = 2, scales = "free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

### Breakdown by polyp location
```{r plot pre vs post by location, fig.height=10}
poloc.df = pre.df %>% filter( Polyp.Loc %in% polyp.locations )

kable( poloc.df %>%
         filter( Protein == "APCS" ) %>%
         group_by( `Risk category`, Polyp.Loc ) %>%
         summarise( Count = n() ) %>% ungroup() %>%
         spread( `Risk category`, Count ), caption = "Counts" )

ggplot( poloc.df, aes( x=Polyp.Loc, y=`Log-measurement`, fill = `Risk category` ) ) +
  facet_wrap( ~Protein, ncol = 2, scales = "free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```
