---
title: "Analysis of pre- and post-polypectomy results"
author: "Yuriy Sverchkov"
date: "4/17/2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("ggplot2")
library("tidyr")
library("knitr")

age.bin.list = c( "Under 50", "50-54", "55-59", "60-64", "65-9", "70+" )
polyp.locations = c( "Colon", "Colon and Rectum", "Rectum" )
```

Comparison of pre- and post- polypectomy data.

## Difference between the pre- and post- polypectomy distributions
```{r prepare data}
summarized.df = readRDS("../data-clean/summarized.v2.rds")

tmp = summarized.df %>%
  mutate( Log.Measurement = log( Geom.Mean ) ) %>%
  filter( grepl( "(Pre|Post)$", Sample.ID ) ) %>%
  mutate(
    Tuple.ID = gsub( "[_-]?(Pre|Post)$", "", Sample.ID ),
    Pre = grepl( "Pre$", Sample.ID ),
    Measurement = ifelse( Pre, "Pre", "Post" ) )

# Extract polup info
polyp.info.df = tmp %>% filter( Pre ) %>% select( Tuple.ID, Protein.Name, Colonoscopy.Pathology, Gender, Age, Age.Bin, Polyp.Loc )
q4.df = tmp %>% select( -Colonoscopy.Pathology, -Gender, -Age, -Age.Bin, -Polyp.Loc ) %>% left_join( polyp.info.df )

proteins = ( summarized.df %>% ungroup() %>% distinct( Protein.Name ) )$Protein.Name
n.proteins = length( proteins )
```

Protein measurements (the logarithm of the geometric mean (over replicates) of area ratios):
```{r plot pre vs post, fig.height=10}
ggplot( q4.df, aes( x="", y=Log.Measurement, fill = Measurement ) ) +
  facet_wrap( ~Protein.Name, ncol = 2, scales="free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

```{r make a dataframe for t-test}
keep = function( which, what ) {
  x = ifelse( which, what, NA )
  if ( all( is.na( x ) ) ) NA
  else x[ !is.na( x ) ]
}

test.df =
  q4.df %>%
  group_by( Tuple.ID, Protein.Name ) %>%
  summarize(
    Pre.Value = keep( Pre, Geom.Mean ),
    Post.Value = keep( !Pre, Geom.Mean ),
    Pre.LogValue = keep( Pre, Log.Measurement ),
    Post.LogValue = keep( !Pre, Log.Measurement ),
    Pre.Risk = keep( Pre, Risk.category.Cancer.stage ),
    Gender = keep( Pre, Gender ),
    Age = keep( Pre, Age ),
    Age.Bin = keep( Pre, Age.Bin ),
    Polyp.Loc = keep( Pre, Polyp.Loc ) ) %>%
  mutate(
    Pre.minus.Post = Pre.Value - Post.Value,
    Pre.to.Post.Ratio = Pre.Value / Post.Value )
```

```{r constants for t-test}
alpha = 0.05
p.value.threshold = alpha/n.proteins
```

For each protein, we use a paired t-test for comparing the results.
Since we test `r n.proteins` proteins, we use a Bonferroni correction and consider only p-values below $`r alpha` / `r n.proteins` = `r p.value.threshold`$ significant.
```{r do t-tests}
t.test.results = Map( function ( protein ){
  local.df = test.df %>% filter( Protein.Name == protein )
  t.test( x = local.df$Pre.Value, y = local.df$Post.Value, paired = TRUE )
}, proteins )

print( "Statistically significant differences found in" )
for ( protein in proteins ) {
  p = t.test.results[[protein]]$p.value
  if( p < p.value.threshold )
    print( paste( protein, "p-value =", p ) )
}
```
Detailed t-test results are in the Supplementary section.

### Breakdown by gender

Protein measurements (the logarithm of the geometric mean (over replicates) of area ratios):
```{r plot pre vs post by gender, fig.height=10}
gender.df = q4.df %>% filter( Gender %in% c("M","F") )

kable( gender.df %>%
         filter( Protein.Name == proteins[1] ) %>%
         group_by( Measurement, Gender ) %>%
         summarise( Count = n() ) %>% ungroup() %>%
         spread( Measurement, Count ), caption = "Counts" )

ggplot( gender.df, aes( x=Gender, y=Log.Measurement, fill = Measurement ) ) +
  facet_wrap( ~Protein.Name, ncol = 2, scales = "free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

### Breakdown by age

```{r plot pre vs post by age, fig.height=10}
age.df = q4.df %>% filter( Age.Bin %in% age.bin.list )

kable( age.df %>%
         filter( Protein.Name == proteins[1] ) %>%
         group_by( Measurement, Age.Bin ) %>%
         summarise( Count = n() ) %>% ungroup() %>%
         spread( Measurement, Count ), caption = "Counts" )

ggplot( age.df, aes( x=Age.Bin, y=Log.Measurement, fill = Measurement ) ) +
  facet_wrap( ~Protein.Name, ncol = 2, scales = "free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

### Breakdown by polyp location
```{r plot pre vs post by location, fig.height=10}
poloc.df = q4.df %>% filter( Polyp.Loc %in% polyp.locations )

kable( poloc.df %>%
         filter( Protein.Name == proteins[1] ) %>%
         group_by( Measurement, Polyp.Loc ) %>%
         summarise( Count = n() ) %>% ungroup() %>%
         spread( Measurement, Count ), caption = "Counts" )

ggplot( poloc.df, aes( x=Polyp.Loc, y=Log.Measurement, fill = Measurement ) ) +
  facet_wrap( ~Protein.Name, ncol = 2, scales = "free" ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

## Differences by risk group
Plot of differences:
```{r plot differences}
ggplot( test.df, aes( x=Protein.Name, y=Pre.minus.Post ) ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

Broken down by pre-polypectomy risk:
```{r plot differences by risk}
ggplot( test.df, aes( x=Protein.Name, y=Pre.minus.Post, fill = Pre.Risk ) ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

Wilcoxon test to determine whether differences in the high-risk group are significantly different from differences in the low-risk group (Bonferroni corrected):
```{r difference wilcox test}
wilcox.difference.results = Map( function (protein) {
  local.df = test.df %>% filter( Protein.Name == protein )
  low.df = local.df %>% filter( Pre.Risk == "Low" )
  high.df = local.df %>% filter( Pre.Risk == "High" )
  wilcox.test( x = low.df$Pre.minus.Post, y = high.df$Pre.minus.Post )
}, proteins)

significant.w.d.r =
  names(
    which(
      Reduce( c, Map(
        function (result) result$p.value < p.value.threshold,
        wilcox.difference.results )
      )
    )
  )

if ( length( significant.w.d.r ) > 0 ){
  print( "Significant differences in:")
  for ( protein in proteins ) {
    p = wilcox.difference.results[[protein]]$p.value
    if( p < p.value.threshold )
      print( paste( protein, "p-value =", p ) )
  }
} else print( "No significant differences found" )
```
Detailed Wilcoxon test results are in the Supplementary section.

## Ratios
Plot of ratios:
```{r plot ratios}
ggplot( test.df, aes( x=Protein.Name, y=Pre.to.Post.Ratio ) ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

Broken down by pre-polypectomy risk:
```{r plot ratios by risk}
ggplot( test.df, aes( x=Protein.Name, y=Pre.to.Post.Ratio, fill = Pre.Risk ) ) +
  geom_boxplot() +
  coord_flip() +
  theme_minimal()
```

Wilcoxon test to determine whether ratios in the high-risk group are significantly different from ratios in the low-risk group (Bonferroni corrected):
```{r ratio wilcox test}
wilcox.ratio.results = Map( function (protein) {
  local.df = test.df %>% filter( Protein.Name == protein )
  low.df = local.df %>% filter( Pre.Risk == "Low" )
  high.df = local.df %>% filter( Pre.Risk == "High" )
  wilcox.test( x = low.df$Pre.to.Post.Ratio, y = high.df$Pre.to.Post.Ratio )
}, proteins)

significant.w.r.r =
  names(
    which(
      Reduce( c, Map(
        function (result) result$p.value < p.value.threshold,
        wilcox.ratio.results )
      )
    )
  )

if ( length( significant.w.r.r ) > 0 ){
  print( "Significant differences in:")
  for ( protein in proteins ) {
    p = wilcox.difference.results[[protein]]$p.value
    if( p < p.value.threshold )
      print( paste( protein, "p-value =", p ) )
  }
} else print( "No significant differences found" )
```
Detailed Wilcoxon test results are in the Supplementary section.

## Supplementary
### Detailed t-test results:
```{r print all t-tests}
t.test.results
```

### Detailed Wilcoxon test results for differences
```{r print all wilcox diff tests}
wilcox.difference.results
```

### Detailed Wilcoxon test results for ratios
```{r print all wilcox ratio tests}
wilcox.ratio.results
```